Spatial variation in predator diel activity patterns -- feral cats avoid red foxes in time, not space

### 2022-04-15 manuscript/diel-manuscript-supp.tex
submission1-JAE

  \journal{?} % Sets Journal name

@@

  \title{Spatial variation in predator diel activity patterns -- feral cats avoid red foxes in time, not space}
    \author[UOM]{Matthew W. Rees\corref{1}}
   \ead{matt.wayne.rees@gmail.com} 
    \author[CEC]{Jack H. Pascoe}

@@

    \author[ARI]{Alan Robley}

@@

    \author[UOM]{Brendan A. Wintle}

@@

      \address[UOM]{Quantitative \& Applied Ecology Group, School of Ecosystem and Forest Science, The University of Melbourne, Parkville, VIC, Australia}
    \address[CEC]{Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia}
    \address[ARI]{Department of Environment, Land, Water and Planning, Arthur Rylah Institute for Environmental Research, Heidelberg, Australia}
      \cortext[1]{Corresponding Author}

@@

\emph{Article type}: Research Article

\emph{Running title}: Spatiotemporal predator activity

\emph{Word count (abstract, body)}: 330, 4338

\emph{References}: ?

\emph{Figures}: 5

\emph{Tables}: 0

\emph{Authorship}:

\emph{Data accessibility}: Data and code will be deposited on the Dryad Digital Repository after acceptance and can be viewed here: \url{https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators}.

@@

Understanding the constraints that apex predators impose on subordinate species is important for anticipating the outcomes of predator management. Subordinate predators may avoid dominant predators in time or space, making it difficult to quantify changes in antipredator behaviours unless joint spatiotemporal analyses are used.

In this study, we tested whether an invasive apex predator (red fox \emph{Vulpes vulpes}) suppresses or alters the spatiotemporal activity of an invasive mesopredator (feral cat \emph{Felis catus}). We surveyed these predators using 3667 camera-trap deployments across two regions of south-eastern Australia; foxes were poison-baited in some landscapes within each region. The simple predator guild in these regions allowed sharp focus on the interactions between these species across experimental gradients of fox (apex predator) activity. We used generalised additive models to quantify overall predator activity across space and fluctuations in predator activity throughout the daily cycle (i.e., diel activity patterns).

When averaged across the study region, red foxes and feral cats had very similar diel activity patterns; however, there was important differentiation at a finer scale. When fox counts at a camera-trap were high, feral cats did not reduce their overall activity but shifted their diel activity patterns to less risky times of the day. In dry habitats of both regions, cats shifted from being nocturnal-crepuscular to mostly diurnal. In wet forest habitat, fox activity was consistent throughout the diel period; but when fox counts were high, cats became more nocturnal, avoiding dawn in particular. Changes in cat diel activity patterns may facilitate spatial coexistence between the two invasive predators, potentially shifting impacts onto different native prey species.

It is well-appreciated that overall predator activity varies spatially and fluctuates throughout the daily cycle. Our study demonstrates that diel activity patterns also vary across space, likely mediated by both landscape context and fear. Apex predator avoidance appears to be dynamic across landscapes of fear---a key nuance which is overlooked when simply comparing the average activity overlap between two species or the spatial overlap of species occurrence.

@@

Predators shape ecosystems through both predation and the fear of predation (Creel \& Christianson 2008; Ritchie \& Johnson 2009). Fear-induced behavioural suppression can be as detrimental to subordinate species as predation itself (Schmitz, Krivan, \& Ovadia 2004; Preisser, Bolnick, \& Benard 2005). These non-consumptive effects of apex predator are expected to be strong drivers of mesopredator behaviour, particularly when resource competition is high (Ritchie \& Johnson 2009). A strengthening or relaxing of antipredator behaviours by mesopredators can have cascading effects across the entire ecosystem: altering population demographics, species interactions, ecological function and human-wildlife coexistence (Brown, Laundré, \& Gurung 1999; Ripple \& Beschta 2004; Estes \emph{et al.} 2011; Gaynor \emph{et al.} 2019; Lamb \emph{et al.} 2020). Hence, understanding how apex predators constrain the behaviour of subordinate species is important to accurately predict the ecosystem-wide consequences of predator management, such as reintroductions or lethal control (Gaynor \emph{et al.} 2021).

Spatial and/or temporal niche partitioning may allow predators to coexist by reducing encounter-rates and resource overlap (Kronfeld-Schor \& Dayan 2003). However, mesopredators may not consistently employ avoidance behaviours because perceived predation risk is temporally and spatially variable, and antipredator behaviours typically involve a trade-off against resource acquisition, such as limiting or relegating activity to suboptimal places or times (Lima \& Dill 1990; Lima \& Bednekoff 1999). Therefore, optimal predator avoidance strategies are likely to vary across heterogeneous landscapes where resource availability (e.g., shelter, food) and perceived predation risks differ (Kauffman \emph{et al.} 2007; Willems \& Hill 2009; Wirsing \emph{et al.} 2021). For example, temporal predator avoidance may be preferable over spatial avoidance if food is constantly available throughout the day, and vice versa. These concepts are unified under the `ecology of fear' (Brown, Laundré, \& Gurung 1999) concept, which has gained increasing attention in recent times (Gaynor \emph{et al.} 2019). Notably, the mesopredator release hypothesis has recently been expanded from increases in mesopredator abundance following apex predator decline (Soulé \emph{et al.} 1988) to also include changes in mesopredator behaviour (Brashares \emph{et al.} 2010).

To accurately quantify avoidance within a predator guild, we first need to understand how the overall activity and diel activity patterns of each species varies `naturally' across landscapes, particularly for species with broad distributions. It is widely recognised that the overall activity of different predator species varies across their distributions, but their diel activity patterns are often assumed to be constant. In this paper, we use the term `overall activity' to refer to the number of `independent' predator detections at a site (offset to account for survey duration; analogous to an activity or abundance index), and `diel activity pattern' to refer to fluctuations in relative activity throughout the 24-hour daily cycle. Overall activity is influenced by predator behaviour, population density and the detection process (Anderson 2001), whereas diel activity patterns are a behavioural trait (less likely to be affected by the detection process, if the survey methodology remains consistent).

Despite modern predator survey technologies providing time-stamped detections, detection times are commonly discarded from analyses, probably because joint modelling of overall activity and diel activity patterns is more complicated. When temporal avoidance is tested, it is usually considered in an ad-hoc fashion, by fitting separate models for spatial and temporal avoidance, or by repeating spatial analyses (e.g., resource selection functions) at different time periods (e.g., Smith \emph{et al.} 2019; Basille \emph{et al.} 2015; Kohl \emph{et al.} 2019). However, discretising the daily cycle into categorical periods (e.g., day and night) introduces bias, assuming animals have complete step-changes in behaviour rather than progressive shifts across the daily cycle. Further, dawn and dusk are particularly important times for many predator species.

Generalised Additive Models (hereafter `GAMs') are increasingly being used to estimate animal diel activity patterns and offer a flexible framework to jointly consider overall activity. GAMs also have other benefits, including smoothing penalties to reduce overfitting, the ability to capture nonlinear interactions between multiple variables with different units, and the ability to share information across categorical variables through hierarchical specifications (Wood 2017; Pedersen \emph{et al.} 2019). However, we are only aware of one study which allowed animal diel activity to interact with predation risk as a continuous variable in a GAM (although without considering overall activity; Cunningham \emph{et al.} 2019).

The red fox \emph{Vulpes vulpes} (hereafter `fox'; \textasciitilde6 kg) and feral cat \emph{Felis catus} (hereafter `cat'; \textasciitilde4 kg) have devastating impacts on native prey throughout their introduced range, implicated in the extinction of \textasciitilde10 and 63 species, respectively (Doherty \emph{et al.} 2016). The impacts of these invasive predators have been particularly extreme on the Australian continent (Woinarski, Burbidge, \& Harrison 2015). Cats are more difficult to manage, and so introduced predator control programs in Australia often target only foxes (particularly through poison-baiting; Reddiex \emph{et al.} 2007). As foxes and cats compete for many of the same resources, there is concern that lethal fox control could cause a mesopredator release (Soulé \emph{et al.} 1988) of feral cats (Glen \& Dickman 2005; Robley \emph{et al.} 2014; Marlow \emph{et al.} 2015; Doherty \& Ritchie 2017; Comer \emph{et al.} 2020; Wayne \emph{et al.} 2017). There is some evidence that feral cats increase in activity (although highly uncertain; Hunter \emph{et al.} 2018), density (Chapter \ref{density}) and alter their diets and use of space (Chapter \ref{density}; Molsher \emph{et al.} 2017) in response to fox control. Other studies have investigated potential spatial and temporal interactions between these invasive predators (e.g., Roshier \& Carter 2021), but not in response to fox control, or in a joint spatiotemporal framework that allows flexibility in cat avoidance behaviours in respect to differences in fox activity.

@@

In the Glenelg region, large patches of natural vegetation are fragmented by pastoral farming and residential properties (Fig. \ref{fig:diel-map}). Foxes in three distinct forest blocks in this region have been subject to poison-baiting since October 2005, with fortnightly bait replacements (Robley \emph{et al.} 2014). These forest blocks, along with three similar, unbaited forest blocks to the north have been simultaneously surveyed annually under the `Glenelg Ark' fox control program since 2005 (40 sites per block; Robley \emph{et al.} 2020). Hair-tubes were used to monitor species from 2005 - 2013 (presented in Robley \emph{et al.} 2014), replaced by camera-traps from 2013; here we present camera-trap data from 2013 - 2019 (Robley \emph{et al.} 2020). We also included a further 425 camera-trap deployments at unique locations from early 2018 (M.W.R. PhD surveys). This totals 2039 camera-trap deployments in the Glenelg region, collected in a control-impact experimental design (foxes had been continuously controlled for at 8 - 14 years in the treatment landscapes at the time of these surveys).

@@

The Otway Ranges is a largely continuous patch of natural vegetation with a strong east-west rainfall gradient (Fig. \ref{fig:diel-map}). A matrix of cool temperate rainforest and wet forest at high altitudes in the south-west descend into a large heathland directly north, and into dry forests and then heathlands to the north-east. Fox-baiting commenced in small sections of the Otway Ranges in 2008 and large-scale systematic baiting began in 2016 - 2017 under the `Otway Ark' program (Robley, Moloney, \& Parks Victoria West Coast District Team 2019). For the first six weeks, poison-baits were replaced weekly, then changing to ongoing monthly bait-replacement. There was a pause in baiting for approximately six months during the second half of 2018. Fox control recommenced in late 2018 with four weeks of fortnightly bait-replacement, before returning to monthly bait-replacement. A large section of the Otway Ranges to the north-west remains unbaited, but is monitored as an experimental non-treatment site (Robley, Moloney, \& Parks Victoria West Coast District Team 2019). Otway Ark managers survey 372 camera-trap sites annually (sequentially across the region); we present one `before' baiting survey and two `after' baiting surveys of each site from 2016 - 2018, totalling 1113 camera-trap deployments (Robley, Moloney, \& Parks Victoria West Coast District Team 2019). We also include data from an additional before-after control-impact surveys (one `before' baiting survey and two `after' bating surveys) in the western section of the Otway Ranges, conducted annually 2017 - 2019 (M.W.R PhD surveys). This added a further 195 sites and 524 camera-trap deployments.

@@

All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (white or infrared flash), attached to a tree or a metal picket, facing a lure. The Glenelg Ark and Otway Ark fox monitoring programs positioned camera-traps at least 40 cm above ground on a tree or a metal picket and angled downwards toward a lure approximately 1 - 1.5 m away (Robley, Moloney, \& Parks Victoria West Coast District Team 2019; Robley \emph{et al.} 2020). The lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake. The M.W.R. PhD surveys across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground) angled only slightly downwards toward a tuna oil lure approximately 2 - 2.5 m away (detailed in Rees \emph{et al.} 2019). Camera-traps were active for an average of 47 days (maximum 93 days), totalling 172,052 trap-nights.

@@

All data analysis was conducted in R version 3.6.3 (R Core Team 2020). We first used lorelograms to identify the minimum interval to approximate independence (Iannarilli \emph{et al.} 2019); this indicated that discarding repeat detections of a species within 30 minutes was sufficient to reduce temporal autocorrelation. To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the `maptools' R-package (Bivand \& Lewin-Koh 2021) and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by Vazquez \emph{et al.} (2019). We then built a dataframe consisting of a row for each hour of the day (0 -- 23), for every camera-trap deployment (n = 3667), recording the total number of `independent' fox and feral cat detections within each hour across the camera-trap survey.

@@

We modelled the total number of independent detections of each predator per hour for each camera-trap deployment (response variable) with generalised additive mixed-effect models implemented in the `mgcv' R-package (Wood 2017). We used the negative binomial family, as overdispersion, but not zero-inflation, was detected with a poisson distribution using the `DHARMa' R-package (Hartig 2020). We specified the natural log of the number of survey days as a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling. For fox models, we also included a smooth effect of poison-bait density with separate responses per region to account for the effect of fox control (all figures in this manuscript are derived from fox models predicted to a no fox-baiting scenario). This formed the base model specification for each model we fitted; models differed in their specification of the cyclical hour smooth to provide inference on variations of predator diel activity across the four questions of interest; this is detailed in the sections below.

@@

To examine how the overall activity and diel activity patterns of each predator varied across space, we fit a model for each predator which included a tensor product interaction between a spatial smooth and hourly smooth. This allowed predators to have different activity levels across space (static across the years surveyed), as well as variation in diel activity pattern across space. Space was modelled using camera-trap coordinates and a duchon spline basis (Miller \& Wood 2014). To examine how the relative strength of diel activity patterns changed across space, we plotted the percentage increase from the minimum to maximum activity estimate within the daily cycle for each predicted location (hereafter referred to as `diel activity pattern strength').

@@

Predator activity varied across space; we hypothesised that this was partly due to differences in vegetation type, based both on the observed spatial patterns and because vegetation type is a major driver of understorey habitat structure and prey occurrence in these regions (Swan \emph{et al.} 2015; Hradsky \emph{et al.} 2017). To test whether the diel activity pattern of each predator varied among vegetation types, we identified the Ecological Vegetation Class group (hereafter `vegetation type'; standard units for vegetation classification in Victoria; Department of Environment, Land, Water \& Planning 2020) for each unique camera-trap site, totalling eight vegetation types. As rainforests are interspersed (primarily in low lying gullies) at fine-scales throughout wet and damp forests in the south-eastern Otway Ranges, we merged them together (hereafter referred to as `wet forests'). We then estimated predator activity across vegetation types using a hierarchical model specification: a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for the seven vegetation types (`model GS' detailed in Pedersen \emph{et al.} 2019). We also included a random effect to account for differences in overall activity levels between the two regions.

@@

A key question in ecological theory is whether animals are evolutionary hardwired to occupy particular temporal niches, or have circadian rhythms that are responsive to changing environmental conditions and interactions with other species (Schoener 1974; Daan 1981; Lima \& Dill 1990). Here we demonstrate that diel activity patterns are not fixed, but vary across space based on landscape context and fear. In our study, sympatric invasive predators had similar diel activity patterns when averaged across broad regions (i.e., high circular overlap; Fig. \ref{fig:diel-space-time-marginal}; as did Roshier \& Carter 2021), but behaviours varied considerably within landscapes. Fox daily activity patterns were most strongly tied to the daily cycle in dry habitat types, but showed little diel activity pattern in the wet forests (Fig. \ref{fig:diel-veg}a). In contrast, cats were mostly nocturnal in wet forests but crepuscular in dry vegetation types (Fig. \ref{fig:diel-veg}b). Within broad habitat types, cats altered their diel activity patterns at sites with higher fox activity (Fig. \ref{fig:diel-cat-fox}). Control programs that reduce invasive fox activity are therefore likely to change cat diel activity patterns, which in turn may alter cat impacts on native prey species. Quantifying changes in diel activity patterns provides important context for understanding species interactions, which is key for effective ecosystem management (Gaynor \emph{et al.} 2021).

Shifting diel activity patterns may facilitate spatial coexistence of dominant and subordinate species (Carothers \& Jaksić 1984). For cats, altering diel activity patterns to less-preferred times of day may be worthwhile to persist in high-quality habitat. Few studies have demonstrated predator-induced shifts in diel activity (Kronfeld-Schor \& Dayan 2003), but notably, ship rats \emph{Rattus norvegicus} were also found to switch from nocturnal to diurnal behaviour in response to fox activity (Fenn \& Macdonald 1995) and a similar nocturnal-diurnal shift was observed in American mink \emph{Neovison vison} following the recolonisation of native predators (Harrington \emph{et al.} 2009). For cats in our study, a switch to diurnal behaviour where fox activity was high in the dry vegetation types may have been facilitated by the higher abundance of reptiles in these habitat types relative to wet forests (which are mostly diurnal; Woinarski \emph{et al.} 2018). At wet forest sites with high fox activity, cats concentrated their activity away from sunrise and sunset towards midnight, despite a diurnal shift appearing to similarly reduce the risk of a fox encounter. In this situation, we expect becoming more nocturnal to be favourable over a diurnal shift because this is when small mammals are active and cats would be least visible to foxes (cats here mostly had black or grey coats). Overall fox activity was twice as low in the wet forests relative to dry habitat types, and so cats were likely under less pressure to radically alter their diel activity patterns. Understanding how these potential avoidance behaviours impacts native prey is a key research priority to improve invasive predator management.

>>>>

  \journal{Journal of Animal Ecology} % Sets Journal name

@@

  \title{Spatial variation in predator diel activity patterns -- feral cats avoid red foxes in time, not space}
    \author[UOM]{Matthew W. Rees\corref{1}}
   \ead{matt.wayne.rees@gmail.com} 
    \author[UOM]{Brendan A. Wintle}

    \author[CEC]{Jack H. Pascoe}

@@

@@

@@

      \address[UOM]{Quantitative \& Applied Ecology Group, School of Ecosystem and Forest Science, The University of Melbourne, Parkville, VIC, Australia}
    \address[CEC]{Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia}
      \cortext[1]{Corresponding Author}

@@

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Understanding the constraints that apex predators impose on subordinate species is important for anticipating the outcomes of predator management. Subordinate predators may avoid dominant predators in time or space, making it difficult to quantify changes in antipredator behaviours unless joint spatiotemporal analyses are used.
\item
  In this study, we tested whether an invasive apex predator (red fox \emph{Vulpes vulpes}) suppresses or alters the spatiotemporal activity of an invasive mesopredator (feral cat \emph{Felis catus}).
\item
  We surveyed these predators using 3667 camera-trap deployments across two regions of south-eastern Australia; foxes were poison-baited in some landscapes within each region. The simple predator guild in these regions allowed sharp focus on the interactions between these species across experimental gradients of fox (apex predator) activity. We used generalised additive models to quantify overall predator activity across space and fluctuations in predator activity throughout the daily cycle (i.e., diel activity patterns).
\item
  When averaged across the study region, red foxes and feral cats had very similar diel activity patterns; however, there was important differentiation at a finer scale. When fox counts at a camera-trap were high, feral cats did not reduce their overall activity but shifted their diel activity patterns to less risky times of the day. In dry habitats of both regions, cats shifted from being nocturnal-crepuscular to mostly diurnal. In wet forest habitat, fox activity was consistent throughout the diel period; but when fox counts were high, cats became more nocturnal, avoiding dawn in particular. Changes in cat diel activity patterns may facilitate spatial coexistence between the two invasive predators, potentially shifting impacts onto different native prey species.
\item
  It is well-appreciated that overall predator activity varies spatially and fluctuates throughout the daily cycle. Our study demonstrates that diel activity patterns also vary across space, likely mediated by both landscape context and fear. Apex predator avoidance appears to be dynamic across landscapes of fear---a key nuance which is overlooked when simply comparing the average activity overlap between two species or the spatial overlap of species occurrence.
\end{enumerate}

@@

Predators shape ecosystems through both predation and the fear of predation (Creel \& Christianson, 2008; Ritchie \& Johnson, 2009). Fear-induced behavioural suppression can be as detrimental to subordinate species as predation itself (Preisser et al., 2005; Schmitz et al., 2004). These non-consumptive effects of apex predator are expected to be strong drivers of mesopredator behaviour, particularly when resource competition is high (Ritchie \& Johnson, 2009). A strengthening or relaxing of antipredator behaviours by mesopredators can have cascading effects across the entire ecosystem: altering population demographics, species interactions, ecological function and human-wildlife coexistence (Brown et al., 1999; Estes et al., 2011; Gaynor et al., 2019; Lamb et al., 2020; Ripple \& Beschta, 2004). Hence, understanding how apex predators constrain the behaviour of subordinate species is important to accurately predict the ecosystem-wide consequences of predator management, such as reintroductions or lethal control (Gaynor et al., 2021).

Spatial and/or temporal niche partitioning may allow predators to coexist by reducing encounter-rates and resource overlap (Kronfeld-Schor \& Dayan, 2003). However, mesopredators may not consistently employ avoidance behaviours because perceived predation risk is temporally and spatially variable, and antipredator behaviours typically involve a trade-off against resource acquisition, such as limiting or relegating activity to suboptimal places or times (Lima \& Bednekoff, 1999; Lima \& Dill, 1990). Therefore, optimal predator avoidance strategies are likely to vary across heterogeneous landscapes where resource availability (e.g., shelter, food) and perceived predation risks differ (Kauffman et al., 2007; Willems \& Hill, 2009; Wirsing et al., 2021). For example, temporal predator avoidance may be preferable over spatial avoidance if food is constantly available throughout the day, and vice versa. These concepts are unified under the `ecology of fear' (Brown et al., 1999) concept, which has gained increasing attention in recent times (Gaynor et al., 2019). Notably, the mesopredator release hypothesis has recently been expanded from increases in mesopredator abundance following apex predator decline (Soulé et al., 1988) to also include changes in mesopredator behaviour (Brashares et al., 2010).

To accurately quantify avoidance within a predator guild, we first need to understand how the overall activity and diel activity patterns of each species varies `naturally' across landscapes, particularly for species with broad distributions. It is widely recognised that the overall activity of different predator species varies across their distributions, but their diel activity patterns are often assumed to be constant. In this paper, we use the term `overall activity' to refer to the number of `independent' predator detections at a site (offset to account for survey duration; analogous to an activity or abundance index), and `diel activity pattern' to refer to fluctuations in relative activity throughout the 24-hour daily cycle. Overall activity is influenced by predator behaviour, population density and the detection process (Anderson, 2001), whereas diel activity patterns are a behavioural trait (less likely to be affected by the detection process, if the survey methodology remains consistent).

Despite modern predator survey technologies providing time-stamped detections, detection times are commonly discarded from analyses, probably because joint modelling of overall activity and diel activity patterns is more complicated. When temporal avoidance is tested, it is usually considered in an ad-hoc fashion, by fitting separate models for spatial and temporal avoidance, or by repeating spatial analyses (e.g., resource selection functions) at different time periods (e.g., Smith et al., 2019; Basille et al., 2015; Kohl et al., 2019). However, discretising the daily cycle into categorical periods (e.g., day and night) introduces bias, assuming animals have complete step-changes in behaviour rather than progressive shifts across the daily cycle. Further, dawn and dusk are particularly important times for many predator species.

Generalised Additive Models (hereafter `GAMs') are increasingly being used to estimate animal diel activity patterns and offer a flexible framework to jointly consider overall activity. GAMs also have other benefits, including smoothing penalties to reduce overfitting, the ability to capture nonlinear interactions between multiple variables with different units, and the ability to share information across categorical variables through hierarchical specifications (Pedersen et al., 2019; Wood, 2017). However, we are only aware of one study which allowed animal diel activity to interact with predation risk as a continuous variable in a GAM (although without considering overall activity; Cunningham et al., 2019).

The red fox \emph{Vulpes vulpes} (hereafter `fox'; \textasciitilde6 kg) and feral cat \emph{Felis catus} (hereafter `cat'; \textasciitilde4 kg) have devastating impacts on native prey throughout their introduced range, implicated in the extinction of \textasciitilde10 and 63 species, respectively (Doherty et al., 2016). The impacts of these invasive predators have been particularly extreme on the Australian continent (Woinarski et al., 2015). Cats are more difficult to manage, and so introduced predator control programs in Australia often target only foxes (particularly through poison-baiting; Reddiex et al., 2007). As foxes and cats compete for many of the same resources, there is concern that lethal fox control could cause a mesopredator release (Soulé et al., 1988) of feral cats (Comer et al., 2020; Doherty \& Ritchie, 2017; Glen \& Dickman, 2005; Marlow et al., 2015; Robley et al., 2014; Wayne et al., 2017). There is some evidence that feral cats increase in activity (although highly uncertain; Hunter et al., 2018), density (M.W.R, unpublished data) and alter their diets and use of space (Molsher et al., 2017) in response to fox control. Other studies have investigated potential spatial and temporal interactions between these invasive predators (e.g., Roshier \& Carter, 2021), but not in response to fox control, or in a joint spatiotemporal framework that allows flexibility in cat avoidance behaviours in respect to differences in fox activity.

@@

In the Glenelg region, large patches of natural vegetation are fragmented by pastoral farming and residential properties (Fig. \ref{fig:diel-map}). Foxes in three distinct forest blocks in this region have been subject to poison-baiting since October 2005, with fortnightly bait replacements (Robley et al., 2014). These forest blocks, along with three similar, unbaited forest blocks to the north have been simultaneously surveyed annually under the `Glenelg Ark' fox control program since 2005 (40 sites per block; Robley et al., 2020). Hair-tubes were used to monitor species from 2005 - 2013 (presented in Robley et al., 2014), replaced by camera-traps from 2013; here we present camera-trap data from 2013 - 2019 (Robley et al., 2020). We also included a further 425 camera-trap deployments at unique locations from early 2018 (M.W.R. PhD surveys). This totals 2039 camera-trap deployments in the Glenelg region, collected in a control-impact experimental design (foxes had been continuously controlled for at 8 - 14 years in the treatment landscapes at the time of these surveys).

@@

The Otway Ranges is a largely continuous patch of natural vegetation with a strong east-west rainfall gradient (Fig. \ref{fig:diel-map}). A matrix of cool temperate rainforest and wet forest at high altitudes in the south-west descend into a large heathland directly north, and into dry forests and then heathlands to the north-east. Fox-baiting commenced in small sections of the Otway Ranges in 2008 and large-scale systematic baiting began in 2016 - 2017 under the `Otway Ark' program (Robley et al., 2019). For the first six weeks, poison-baits were replaced weekly, then changing to ongoing monthly bait-replacement. There was a pause in baiting for approximately six months during the second half of 2018. Fox control recommenced in late 2018 with four weeks of fortnightly bait-replacement, before returning to monthly bait-replacement. A large section of the Otway Ranges to the north-west remains unbaited, but is monitored as an experimental non-treatment site (Robley et al., 2019). Otway Ark managers survey 372 camera-trap sites annually (sequentially across the region); we present one `before' baiting survey and two `after' baiting surveys of each site from 2016 - 2018, totalling 1113 camera-trap deployments (Robley et al., 2019). We also include data from an additional before-after control-impact surveys (one `before' baiting survey and two `after' bating surveys) in the western section of the Otway Ranges, conducted annually 2017 - 2019 (M.W.R PhD surveys). This added a further 195 sites and 524 camera-trap deployments.

@@

All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (white or infrared flash), attached to a tree or a metal picket, facing a lure. The Glenelg Ark and Otway Ark fox monitoring programs positioned camera-traps at least 40 cm above ground on a tree or a metal picket and angled downwards toward a lure approximately 1 - 1.5 m away (Robley et al., 2019, 2020). The lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake. The M.W.R. PhD surveys across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground) angled only slightly downwards toward a tuna oil lure approximately 2 - 2.5 m away (detailed in Rees et al., 2019). Camera-traps were active for an average of 47 days (maximum 93 days), totalling 172,052 trap-nights.

@@

All data analysis was conducted in R version 3.6.3 (R Core Team, 2020). We first used lorelograms to identify the minimum interval to approximate independence (Iannarilli et al., 2019); this indicated that discarding repeat detections of a species within 30 minutes was sufficient to reduce temporal autocorrelation. To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the `maptools' R-package (Bivand \& Lewin-Koh, 2021) and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by Vazquez et al. (2019). We then built a dataframe consisting of a row for each hour of the day (0 -- 23), for every camera-trap deployment (n = 3667), recording the total number of `independent' fox and feral cat detections within each hour across the camera-trap survey.

@@

We modelled the total number of independent detections of each predator per hour for each camera-trap deployment (response variable) with generalised additive mixed-effect models implemented in the `mgcv' R-package (Wood, 2017). We used the negative binomial family, as overdispersion, but not zero-inflation, was detected with a poisson distribution using the `DHARMa' R-package (Hartig, 2020). We specified the natural log of the number of survey days as a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling. For fox models, we also included a smooth effect of poison-bait density with separate responses per region to account for the effect of fox control (all figures in this manuscript are derived from fox models predicted to a no fox-baiting scenario). This formed the base model specification for each model we fitted; models differed in their specification of the cyclical hour smooth to provide inference on variations of predator diel activity across the four questions of interest; this is detailed in the sections below.

@@

To examine how the overall activity and diel activity patterns of each predator varied across space, we fit a model for each predator which included a tensor product interaction between a spatial smooth and hourly smooth. This allowed predators to have different activity levels across space (static across the years surveyed), as well as variation in diel activity pattern across space. Space was modelled using camera-trap coordinates and a duchon spline basis (Miller \& Wood, 2014). To examine how the relative strength of diel activity patterns changed across space, we plotted the percentage increase from the minimum to maximum activity estimate within the daily cycle for each predicted location (hereafter referred to as `diel activity pattern strength').

@@

Predator activity varied across space; we hypothesised that this was partly due to differences in vegetation type, based both on the observed spatial patterns and because vegetation type is a major driver of understorey habitat structure and prey occurrence in these regions (Hradsky et al., 2017; Swan et al., 2015). To test whether the diel activity pattern of each predator varied among vegetation types, we identified the Ecological Vegetation Class group (hereafter ``vegetation type''; standard units for vegetation classification in Victoria; Department of Environment, Land, Water \& Planning, 2020) for each unique camera-trap site, totalling eight vegetation types. As rainforests are interspersed (primarily in low lying gullies) at fine-scales throughout wet and damp forests in the south-eastern Otway Ranges, we merged them together (hereafter referred to as `wet forests'). We then estimated predator activity across vegetation types using a hierarchical model specification: a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for the seven vegetation types (``model GS'' detailed in Pedersen et al., 2019). We also included a random effect to account for differences in overall activity levels between the two regions.

@@

A key question in ecological theory is whether animals are evolutionary hardwired to occupy particular temporal niches, or have circadian rhythms that are responsive to changing environmental conditions and interactions with other species (Daan, 1981; Lima \& Dill, 1990; Schoener, 1974). Here we demonstrate that diel activity patterns are not fixed, but vary across space based on landscape context and fear. In our study, sympatric invasive predators had similar diel activity patterns when averaged across broad regions (i.e., high circular overlap; Fig. \ref{fig:diel-space-time-marginal}; as did Roshier \& Carter, 2021), but behaviours varied considerably within landscapes. Fox daily activity patterns were most strongly tied to the daily cycle in dry habitat types, but showed little diel activity pattern in the wet forests (Fig. \ref{fig:diel-veg}a). In contrast, cats were mostly nocturnal in wet forests but crepuscular in dry vegetation types (Fig. \ref{fig:diel-veg}b). Within broad habitat types, cats altered their diel activity patterns at sites with higher fox activity (Fig. \ref{fig:diel-cat-fox}). Control programs that reduce invasive fox activity are therefore likely to change cat diel activity patterns, which in turn may alter cat impacts on native prey species. Quantifying changes in diel activity patterns provides important context for understanding species interactions, which is key for effective ecosystem management (Gaynor et al., 2021).

Shifting diel activity patterns may facilitate spatial coexistence of dominant and subordinate species (Carothers \& Jaksić, 1984). For cats, altering diel activity patterns to less-preferred times of day may be worthwhile to persist in high-quality habitat. Few studies have demonstrated predator-induced shifts in diel activity (Kronfeld-Schor \& Dayan, 2003), but notably, ship rats \emph{Rattus norvegicus} were also found to switch from nocturnal to diurnal behaviour in response to fox activity (Fenn \& Macdonald, 1995) and a similar nocturnal-diurnal shift was observed in American mink \emph{Neovison vison} following the recolonisation of native predators (Harrington et al., 2009). For cats in our study, a switch to diurnal behaviour where fox activity was high in the dry vegetation types may have been facilitated by the higher abundance of reptiles in these habitat types relative to wet forests (which are mostly diurnal; Woinarski et al., 2018). At wet forest sites with high fox activity, cats concentrated their activity away from sunrise and sunset towards midnight, despite a diurnal shift appearing to similarly reduce the risk of a fox encounter. In this situation, we expect becoming more nocturnal to be favourable over a diurnal shift because this is when small mammals are active and cats would be least visible to foxes (cats here mostly had black or grey coats). Overall fox activity was twice as low in the wet forests relative to dry habitat types, and so cats were likely under less pressure to radically alter their diel activity patterns. Understanding how these potential avoidance behaviours impacts native prey is a key research priority to improve invasive predator management.

Cats may have avoided foxes in time, but we saw no sign of spatial avoidance (i.e.~no fine-scale negative association between fox and cat overall activity (Fig. \ref{fig:diel-cat-fox}b:c). Overlap in spatial activity at fine scales (i.e., at the level of camera-trap sites) may have been facilitated by temporal avoidance (Kronfeld-Schor \& Dayan, 2003). However, we considered spatial avoidance across the entire survey duration (averaging 47 days). Cats may indeed avoid foxes spatially, but transiently on the scale of hours to days---after all, how is a cat to know where to avoid a fox without encountering signs of one? Short-term spatial avoidance is quite plausible given foxes mark territories using scats and odours, which cats could tangibly associate with high risk shortly after. Temporary spatial avoidance could be tested using decay curves (e.g., Niedballa et al., 2019). Alternatively, no sign of spatial avoidance between these invasive predators could also be an artefact of the quality of camera set-up and hence detectability. The 3667 camera-traps were deployed by numerous people, and the quality of set-ups differed considerably in terms of detecting predators. Camera-traps that angled only slightly downwards, rather than upwards or strongly downwards, seemed most effective at detecting both predator species (M.W Rees, personal observation). Predator interactions are routinely inferred through spatial associations between species; however such analyses are subject to numerous pitfalls which can make inference unreliable (reviewed in Blanchet et al., 2020).
